PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2022-09-01 manuscript_00.Rmd
freeze 01 release

data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))

@@

@@

```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap}

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

Genes which have little existing knowledge can be considered

Of course other possibilities for ranking genes are possible. We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization. The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.

@@

consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide,
                                           column_labels = consist_diff_organism,
                                           column_split = consist_diff_cts,
                                           row_split = consist_diff_base,
                                           show_row_dend = FALSE,
                                           column_title_rot = 90,
                                           name = 'log2FC', use_raster = TRUE,
                                           row_title_rot = 0) +
  ComplexHeatmap::Heatmap(as.character(query0),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          show_heatmap_legend = FALSE,
                          column_labels = "Pubmed Gene Citations",
                          column_names_side = 'top') +
  ComplexHeatmap::Heatmap(as.character(query2),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          name = 'Citation Count',
                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                          column_names_side = 'top')
ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")

#toc()
```

@@

num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
sup_scEiaD_data_overview_cap <- supFig_cap(name = '_supscEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD. A. tabulation of the number of cells present across {num1} ocular and {num2} non-ocular tissues, B. Number of studies for each ocular tissue. C. Number of cells present across {num3} curated ocular-derived cell types. D. Proportion of cell types (predicted) across each study (split by species and tissue)."))

@@

```{r undersupported_genes, fig.height=48, fig.width=8, echo = FALSE}
cone_undersupported <- queries %>% filter(query2 <= 1) %>% arrange(-`mean log2FC`) %>% filter(!grepl("Endo|Macro|Micro|Endo", Base)) %>% group_by(Symbol) %>% summarise(Count = n(), Base = paste0(Base, collapse = ', '), l2FC = mean(`mean log2FC`)) %>% filter(Count == 1) %>% filter(Base == 'Rod') %>% arrange(-l2FC)

input <- list()
input$exp_plot_groups <- c('study_accession')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- scEiaD_2020_v01 %>% tbl('Genes') %>% collect() %>% mutate(Symbol = str_extract(Gene, '.* \\(') %>% gsub(' \\(','',.)) %>% filter(Symbol %in% cone_undersupported$Symbol) %>% pull(Gene)
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50
input$flip_facet <- FALSE
input$figure <- FALSE
input$figure_bloop <- TRUE
input$figure_fovea <- FALSE

make_exp_plot(input, scEiaD_2020_v01, meta_filter)

```
```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}
PLAE_screenshots_supplemental_cap <- supFig_cap(name='PLAE_screenshots_supplemental', caption = 'a. Recreation of Figure 4 directly in the plae.nei.nih.gov web app. b. Identify potential bipolar cell markers with the precomputed differential testing and in c. visualize the top five genes with a Heatmap.')
knitr::include_graphics('fig/plae_sup_fig_screenshots.png')

>>>>

#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))

@@

## PubMed Citation Search

The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function. We then used the returned Entrez identifier to pull the abstract with the fetch_pubmed_data function so we could extract the PubMed ID (PMID). This was repeated with the the genes identified as being well supported differentially expressed for ocular cell types. To assess whether the difference in distributions was statistically different, we used the base R t.test function.

@@

```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap, echo = FALSE}

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

Genes with few citations can be considered as novel markers. We see, for example, Onecut3 appears to well separate the horizontal cells and has few PubMed citations. To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`). Most of these genes are differentially expressed across more than one cell type. We hand selected C1ql2 and Cartpt (amacrine), Frzb and Slitrk2 (Muller glia), and Onecut3 (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`). All five look to be specific to a cell type, except for Frzb which also is expressed in the RPE.

Other possibilities for ranking genes are possible. We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how one can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization. The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.

@@

consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide,
                                           column_labels = consist_diff_organism,
                                           column_split = consist_diff_cts,
                                           row_split = consist_diff_base,
                                           show_row_dend = FALSE,
                                           column_title_rot = 90,
                                           name = 'log2FC', use_raster = TRUE,
                                           row_title_rot = 0) +
  ComplexHeatmap::Heatmap(as.character(query0),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          show_heatmap_legend = FALSE,
                          column_labels = "Pubmed Gene Citations",
                          column_names_side = 'top') +
  ComplexHeatmap::Heatmap(as.character(query2),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          name = 'Citation Count',
                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                          column_names_side = 'top')
ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
queries %>% select(Symbol, Base, `mean log2FC`, `mean padj`, pm_query0, query0, pm_query2, query2) %>% write_csv("supplemental_files/consistently_differentially_expressed_diff_table.csv.gz")
#toc()
```

@@

num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
sup_scEiaD_data_overview_cap <- supFig_cap(name = 'sup_scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD. A. tabulation of the number of cells present across {num1} ocular and {num2} non-ocular tissues, B. Number of studies for each ocular tissue. C. Number of cells present across {num3} curated ocular-derived cell types. D. Proportion of cell types (predicted) across each study (split by species and tissue)."))

@@

```{r undersupported_genes, fig.height=23, fig.width=8, echo = FALSE, fig.cap = undersupported_genes_cap}
undersupported_genes_cap <- supFig_cap(name = 'undersupported_genes', caption = 'Well supported genes with a mean log2FC > 5 and zero pubmed citation hits for [gene] AND Retina search.')

g_lightlyCited <- queries %>%  filter(`mean log2FC` > 5, query2 < 1) %>% filter(Gene %in% top_genes$Gene) %>% group_by(Base) %>% top_n(12, `mean log2FC`) %>% pull(Gene)

consist_diff_lightlyCited_long <- scEiaD_2020_v01 %>%
  tbl("diff_testing") %>%
  filter(Gene %in% g_lightlyCited) %>%
  collect() %>%
  filter(Base %in% top_genes$Base,
         !Base %in% c('Early RPC','Late RPC'), 
         Group == 'CellType (Predict)',
         Against == 'All') %>%
  select(Gene, Organism, log2FoldChange, Base)

consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_long %>% 
  mutate(GO = paste0(Base, '_', Organism)) %>% 
  select(-Organism, -Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = GO)
consist_diff_lightlyCited_celltype <- colnames(consist_diff_lightlyCited_wide[,-1])
consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_wide %>% data.frame()
row.names(consist_diff_lightlyCited_wide) <- consist_diff_lightlyCited_wide$Gene

consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_wide[,-1]

consist_diff_lightlyCited_organism <- str_extract(consist_diff_lightlyCited_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
consist_diff_lightlyCited_cts  <- str_extract(consist_diff_lightlyCited_celltype, '.*_') %>% gsub('_','',.)
consist_diff_lightlyCited_genes <- row.names(consist_diff_lightlyCited_wide) %>% gsub(' \\(.*','',.)

consist_diff_lightlyCited_base <- consist_diff_lightlyCited_genes %>% as_tibble() %>% 
  rename(Symbol = value) %>% 
  left_join(top_genes %>% 
              mutate(Symbol = gsub(' \\(.*','',Gene)) %>% 
              select(Symbol, Base) %>% 
              unique()) %>% 
  pull(Base)

query0_lightlyCited <- cbind(consist_diff_lightlyCited_genes, consist_diff_lightlyCited_base) %>% as_tibble() %>%
  left_join(queries, by = c('consist_diff_lightlyCited_genes' = 'Symbol', 'consist_diff_lightlyCited_base' = 'Base')) %>%
  mutate(citation = case_when(
    query0 > mean(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
    query0 > median(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
    TRUE ~ "Few")) %>%
  pull(citation)

# pmid2 and pmid3 is "AND RETINA"
# (t.test(pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist(), pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()))$p.value
query2_lightlyCited <- cbind(consist_diff_lightlyCited_genes, consist_diff_lightlyCited_base) %>%
  as_tibble() %>%
  left_join(queries, by = c('consist_diff_lightlyCited_genes' = 'Symbol', 'consist_diff_lightlyCited_base' = 'Base')) %>%
  mutate(citation = case_when(
    query2 > mean(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
    query2 > median(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
    TRUE ~ "Few")) %>%
  pull(citation)
#rand_col <-  circlize::colorRamp2(c('0', '1', '2'), c("white", "gray", "black"))
citation_col <-c('white','gray','black')
names(citation_col) <-  c('Few','Some','Many')

consist_diff_lightlyCited_hm <- ComplexHeatmap::Heatmap(consist_diff_lightlyCited_wide,
                                                        column_labels = consist_diff_lightlyCited_organism,
                                                        column_split = consist_diff_lightlyCited_cts,
                                                        row_split = consist_diff_lightlyCited_base,
                                                        show_row_dend = FALSE,
                                                        column_title_rot = 90,
                                                        name = 'log2FC', use_raster = TRUE,
                                                        row_title_rot = 0) +
  ComplexHeatmap::Heatmap(as.character(query0_lightlyCited),
                          row_split = consist_diff_lightlyCited_base,
                          col = citation_col,
                          show_heatmap_legend = TRUE,
                          column_labels = "Pubmed Gene Citations",
                          column_names_side = 'top') +
  ComplexHeatmap::Heatmap(as.character(query2_lightlyCited),
                          row_split = consist_diff_lightlyCited_base,
                          col = citation_col,
                          show_heatmap_legend = FALSE,
                          row_labels = consist_diff_lightlyCited_genes,
                          name = 'Citation Count',
                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                          column_names_side = 'top')
ComplexHeatmap::draw(consist_diff_lightlyCited_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")

```{r lightlyCited_candidates, echo = F, message = F, fig.cap = lightlyCited_candidates_cap}
lightlyCited_candidates_cap <- supFig_cap(name='lightlyCited_candidates', caption = 'Five hand-selected genes selected from Supplementary Figure X as having fairly unique expression to a celltype while having few pubmed citations. Plotted as expression/boxplots to confirm specificity at the study level.')
knitr::include_graphics('fig/lightlyCited_exp_plot.png')

```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}
PLAE_screenshots_supplemental_cap <- supFig_cap(name='PLAE_screenshots_supplemental', caption = 'a. Recreation of Figure 4 directly in the plae.nei.nih.gov web app. b. Identify potential bipolar cell markers with the precomputed differential testing and in c. visualize the top five genes with a Heatmap.')
knitr::include_graphics('fig/plae_sup_fig_screenshots.png')
